---
title: dt04_regress_onto_SAM.qmd
abstract: |
  The purpose of this notebook is to regress RACMO variables onto the SAM index.
  It is also possible to regress ERA5 variables onto the SAM index as well.
  
  The regressions are run in this notebook (each takes >2-5 minutes) and saved 
  as NetCDFs; they can then be plotted using the an03 notebook.
  
  Each NetCDF output has three layers:
    - regression :      the amount of change in a variable for a unit decrease
                        in the SAM index;
    - pValues    :      the statistical significance of the regression;
    - rSquared   :      the coefficient of determination (i.e. how much of the
                        RACMO/ERA5 variability is explained by the SAM variability).
                        
  The following RACMO variables can be used:
    - snowmelt, albd, t2m;
    - seb, radi, gbot;
    - turb, latf, senf;
    - swsd, swsu, swsn; 
    - lwsd, lwsu, lwsn;
    - w10m, u10m, v10m;
    - smb, precip, sndiv, subl.
    
  The following variables can only be used if they were already created in dt01:
    - turb, radi, w10m, seb, albd, swsn, lwsn
  
  The following ERA5 variables can also be used (if downloaded from CDS); these 
  are very slow! (> 25 min)
    - MSLP, z700
    
  It is possible to choose which SAM index to use: either the NOAA CPC index 
  (set as "noaa"), or the Marshall (2003) index (set as "marshall"). Check (or 
  redefine) the file paths that are expected, in "R/align_sam_to_summers.R".
  
  It is possible to run the regressions for:
    - a single month 
        - e.g. regress the Dec precipitation onto the Dec SAM
    - a multi-month value: 
        - e.g. regress the mean DJ t2m onto the mean DJ SAM
        - e.g. regress the total DJ snowmelt onto the mean DJ SAM
  
  There is a conversion from Joules to Watts which assumes that all of the 
  included months have 31 days. The code for this part in the "Convert units" 
  chunk (line 124). Any changes to account for non-31 day months will need to be 
  hard-coded in.
  
params:
  regressVar: "snowmelt"
  months:     [12, 1]
  samIndex:   "noaa"
  summers:    !expr c(1980:2018)
---

# Data Prep & Set-Up ===========================================================
```{r}
## Run set-up scripts ----------------------------------------------------------
source("R/su01_set_up.R")
```

## Access SAM index
```{r}
samSummer <- align_sam_to_summers(filePath = params$samIndex) |>
  domR::sift("Summer in vector", params$summers)
samMean <- samSummer[month.abb[params$months]] |> 
  rowMeans() |> unlist()        # handles multi-month average
```

## Access RACMO/ERA5 Data
```{r}
if (params$regressVar %in% pToken$varERA5) {
  ### ERA5 Data
  era5Path <- "../../Data/ERA5/Antarctic_DJ_variables/"
  varPath  <- paste0(era5Path, 
                     "ERA5_", toupper(params$regressVar), 
                     "_1979-2021_DJ_monthly.nc")
  varData <- terra::rast(varPath) |>
    terrapin::subset_by_summer(params$summers) |>
    terrapin::subset_by_month(params$months, excludeIncomplete = 3)
  
  # Assign crs as it is not read correctly
  terra::crs(varData) <- terra::crs("epsg:4326")
  
  # Only part of the southern hemisphere
  varData <- terra::crop(varData, terra::ext(c(-180, 180, -90, -25)))
} else if (params$regressVar %in% pToken$varRACMO) {
  ### Read RACMO data
  varData <- polarcm::subset_racmoM_by_summer(racmoData = params$regressVar, 
                                              summers   = params$summers, 
                                              australSplit = 3) |>
    polarcm::subset_racmoM_by_month(months = params$months, 
                                    excludeIncomplete = 3)
} else {
  stop("Unknown variable!")
}
```

### Prep data
Run the remaining chunks regardless of using RACMO or ERA5 data. For some 
variables it makes sense to think of multi-month data as the mean value (e.g. 
the average of the Dec and Jan t2m), whereas others make more sense as totals 
(e.g. the sum of the Dec and Jan snowmelt).
```{r}
# Calculate a mean if necessary
if (params$regressVar %in% pToken$varMeanable) {
  varAustral <- polarcm::calc_fun_australM(x = varData, 
                                           FUN = "mean", 
                                           months = params$months, 
                                           australSplit = 3)
} else if (params$regressVar %in% pToken$varSummable) {
  varAustral <- polarcm::calc_fun_australM(x = varData, 
                                           FUN = "sum", 
                                           months = params$months, 
                                           australSplit = 3)
} else {
  stop("Unknown variable!")
}

varInfo <- list()
varInfo$units <- terra::units(varAustral)      # we'll add this back later
```

## Convert units
```{r}
# from Joules to Watts
if (params$regressVar %in% pToken$varEnergy) {
  varAustral <- varAustral / (60 * 60 * 24 * 31)  # !! assumes Jan / Dec
  varInfo$units <- gsub("J m-2", "W m-2", varInfo$units)
}

# from Kelvin to Celsius
if (params$regressVar %in% pToken$varTemperature) {
  varAustral <- varAustral - 273.15
  varInfo$units <- gsub("K", "ºC", varInfo$units)
}

# from geopotential to geopotential heights
if (params$regressVar %in% pToken$varGeopotential) {
  varAustral <- varAustral / 9.8065
  varInfo$units <- gsub("m**2 s**-2", "m", varInfo$units)
}

# from Pascals to hectoPascals
if (params$regressVar %in% pToken$varPressure) {
  varAustral <- varAustral / 100
  varInfo$units <- gsub("Pa", "hPa", varInfo$units)
}
```

# Perform Regression ===========================================================
This can be a little bit slow; expect it to take at least 2-5 minutes.
```{r}
# Verify the number of summers is the same in the variable and the SAM data
if (length(samMean) != terra::nlyr(varAustral)) stop("Something went wrong!")

# Regression
start <- domR::start_timer(params$regressVar, "regression")
regression <- terra::app(x = varAustral,   # Calc. the change in this variable
                         xx = -samMean,    #    per unit *decrease* in the SAM;
                         detrend = TRUE,   # Linearly detrend both time series
                         fun = apply_lm)   # Use local function in R/ directory
end <- domR::end_timer(start, params$regressVar, "regression")
domR::cat_list(params)

# Naming
rMonths <- domR::get_initials(month.abb[params$months])  # so DJ, not c(12, 1)
names(regression) <- paste(params$regressVar,
                           rep(rMonths, 3),
                           c("regression", "pValues", "rSquared"),
                           sep = "_")
```

# Save =========================================================================
!! Reproject any ERA5 data (EPSG:4326) to RACMO projection?
Makes it quicker/easier to plot in a for-loop in the an03 notebook.
Ignore the warning, GDAL just has issues with the RACMO projection.
```{r}
if (params$regressVar %in% pToken$varERA5) {
  regression <- terra::project(regression, polarcm::use_crs("racmo"))
}
```

## Complete file ready for saving
```{r}
toSave <- terra::sds(list(c(regression[[1]] |> round(3)),   # regression
                          c(regression[[2]] |> round(3)),   # pValues
                          c(regression[[3]] |> round(3))))  # rSquared

# Add units, longname and varname metadata
if (params$regressVar %in% pToken$varMeanable) {
  terra::units(toSave)[1] <- paste(varInfo$units[1], "per unit SAM-")
} else if (params$regressVar %in% pToken$varSummable) {
  terra::units(toSave)[1] <- paste(varInfo$units[1],  "per unit SAM-")
}
terra::units(toSave)[2:3] <- c("p-value", "r2")
terra::varnames(toSave) <- c("regression", "pValue", "rSquared")
terra::longnames(toSave) <- c(paste("Regression of", params$regressVar,
                                    "onto the", params$samIndex, "SAM index.",
                                    "See units."),
                             "Statistical Significance",
                             "Coefficient of Determination")
blurb <- paste("Regression of", params$regressVar, "onto the", params$samIndex, 
               "SAM index, using", rMonths, "data (", min(params$summers), "-",
               max(params$summers), "). Variables are taken from RACMO2.3p3",
               "(van Dalum et al., 2022) or ERA5 (Hersbach et al., 2020), at", 
               "monthly resolution. For temperatures, albedo, wind, and",
               "pressures, the regression is based on the mean of the monthly",
               "values each summer (e.g. average of the Dec and Jan t2m); for",
               "the other variables (i.e. masses & fluxes), the regression is ",
               "based on the sum of the monthly values each summer (e.g. total",
               "of the Dec and Jan snowmelt). The SAM values are the mean of",
               "the monthly values each summer (e.g. average of the Dec and",
               "Jan SAM value).")
```

## Save as a NetCDF
```{r}
# Name set-up
dirName <- domR::cat_path("Data/samRegressions", 
                          paste(params$samIndex, "SAM", rMonths, sep = "_"),
                          paste0("regress", 
                                 paste(range(params$summers), collapse = "_")))

# Create dir if necessary
if (!dir.exists(dirName)) dir.create(dirName, recursive = TRUE)

fileName <- dirName |>
  paste0("regress") |>
  paste(params$regressVar, params$samIndex, rMonths, sep = "_") |>
  paste0("-dt04.nc")

terra::writeCDF(toSave,
                filename = fileName, overwrite = TRUE,
                atts = domR::get_metadata(format = "NetCDF",
                                          blurb = blurb))  
domR::print_on_saving(fileName)
```

# Check output
```{r}
tt <- terra::rast(fileName)
tt
terra::plot(tt$regression, range = c(0.5, 5.5), col = kulaS(10))
```
